The socialization of meritocracy and market justice preferences at school

### 2024-08-05 papelera/paper.tex
Actualización paper anonimizado en .tex y orden de repo

% TODO: Add custom LaTeX header directives here
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The socialization of meritocracy and market justice preferences at school},
  % pdfauthor={Juan Carlos Castillo; Mauricio Salgado; Kevin Carrasco; Andreas Laffert},
  pdfkeywords={market
justice, meritocracy, socialization, family, schools, Chile},
  pdfcreator={LaTeX via pandoc},
  unicode=true,
	bookmarksopen={true},
	pdffitwindow=true, 
	colorlinks=true, 
	linkcolor=bluecite, 
	citecolor=bluecite, 
	urlcolor=bluecite, 
	hyperfootnotes=true, 
	pdfstartview={FitH},
	pdfpagemode= UseNone}

@@

% Author Orchid ID: enter ID or remove command
%\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name
\newcommand{\orcidauthorA}{0000-0003-1265-7854} 
\newcommand{\orcidauthorB}{0000-0002-8108-6154} 
\newcommand{\orcidauthorC}{0000-0002-9348-8361} 
\newcommand{\orcidauthorD}{0000-0002-9008-2454}

@@

%\Author{Juan Carlos Castillo$^{1}$\orcidA{}, 
%Mauricio Salgado$^{2}$\orcidB{}, 
%Kevin Carrasco$^{3}$\orcidC{}~and~Andreas Laffert$^{4}$\orcidD{}
%}
%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
%\AuthorNames{
%  Juan Carlos Castillo,
%  Mauricio Salgado,
%  Kevin Carrasco,
%  Andreas Laffert}

% MDPI internal command: Authors, for citation in the left column
%\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
%\usepackage{xstring}
%\newcommand\getfirst[1]{\getfirstaux#1\relax} \def\getfirstaux#1#2\relax{#1}
%\newcommand{\authorformatted}{Castillo, \getfirst{Juan Carlos}.; Salgado, \getfirst{Mauricio}.; Carrasco, \getfirst{Kevin}.; Laffert, \getfirst{Andreas}.}
%\AuthorCitation{\authorformatted}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
%\address {%
%$^{1}$ \quad Department of Sociology, Universidad de Chile\\
%$^{2}$ \quad CEP, Centro de Estudios Públicos\\
%$^{3}$ \quad COES, Center for Social Conflict and Cohesion Studies\\
%$^{4}$ \quad Institute of Sociology, Pontificia Universidad Católica de Chile\\
%}

@@

% Contact information of the corresponding author
%% see: https://quarto.org/docs/journals/authors.html
%\corres{Correspondence: juancastillov@uchile.cl}

@@

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
\citep[p.384]{lane_market_1986}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
\citep{greitemeyer_subjective_2016, mishra_subjective_2015, smith_relative_2012, power_deprivationprotest_2018},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see \citep{wilson_role_2003}).

@@

Attending to the socialization within the family, the classic work of
Kohn showed that middle-class parents value the expression of internal
states and emotions, such as self-control, curiosity, happiness, and
consideration, while working-class parents promote deference, obedience,
and conformity to authority \citep{kohn_social_1963, kohn_class_1969}.
Although parents from all social backgrounds encourage individualism in
their children, this shared norm translates into different forms in high
and low social classes \citep{kohn_class_1969}.
\citet{acemoglu_obedience_2021} claimed that the values families impart
to their children interact with social mobility. On the one hand,
children from privileged families are socialized to adopt a clear
conception of individualism that highlights their internal states,
independence, and idiosyncrasies. In contrast, children from
disadvantaged families are socialized to support a more balanced view of
individualism that considers personal characteristics as resources to
overcome collective impediments on the path to upward mobility
\citep{iacoviello_collectivism_2019}. In the same line,
\citet{almas_fairness_2017} found that adolescents from
low-socioeconomic-status families are likelier to have an egalitarian
fairness view and consider an equal distribution as fair in a situation
with unequal merits. Taking this into account, we believe that there
will be differences in the socialization of values according to
socioeconomic differences in families that could influence market
justice preferences, which leads us to the following hypothesis:

@@

\includegraphics{template-mdpi_files/figure-pdf/fig-marketjustice-1.pdf}

@@

\includegraphics{template-mdpi_files/figure-pdf/fig-meritocracy-1.pdf}

@@

\includegraphics{template-mdpi_files/figure-pdf/fig-bivariate-1.pdf}

@@

\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabu} to \linewidth {>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}}
\toprule
\multicolumn{1}{c}{\textbf{ }} & \multicolumn{5}{c}{\textbf{Market Justice Preferences}} \\
\cmidrule(l{3pt}r{3pt}){2-6}
Variable & School talent & School effort & Social talent & Social effort & Deservingness\\
\midrule
\textbf{University or Postgraduate} & -0.034 & -0.012 & -0.073. & -0.079* & -0.118**\\
\cmidrule{1-6}
\textbf{More than 25 books} & -0.015 & -0.046 & -0.029 & -0.081** & -0.035\\
\cmidrule{1-6}
\textbf{School SES high} & 0.004 & 0.001 & -0.057. & -0.063* & -0.08**\\
\cmidrule{1-6}
\textbf{Simce Medium} & 0.001 & 0.032 & -0.037 & -0.041 & -0.082*\\
\cmidrule{1-6}
\textbf{Simce High} & -0.016 & 0.008 & -0.043 & -0.063. & -0.087*\\
\bottomrule
\multicolumn{6}{l}{\textsuperscript{} Note: ***p < 0.001, **p < 0.01, *p < 0.05. Source: own elaboration based on First Study of Civic}\\
\multicolumn{6}{l}{Education 2017.}\\
\end{tabu}}

@@

\includegraphics{template-mdpi_files/figure-pdf/fig-interaction-1.pdf}

@@

Our research aimed to examine the relationship between the perception of
meritocracy and the justification of market justice preferences among
eighth-grade students in Chile. From our knowledge this is the first
study addressing market justice preferences at school level, finding
that students exhibit large preferences for market justice (about one
third agree or strongly agree with them), a share that is quite larger
when compared with evidence in adult population. Perception of
meritocracy is also high, particularly when it comes to the reward of
effort, with an striking eighty six percent of agreement. Regarding
these descriptive indicators, a a first message coming from this study
is that there is something particular in school-age when it comes to
perceptions and preferences about inequality. This can relate to
development aspects
\citep{rizzo_children_2020, kim_socioeconomic_2020} but it could
be something specific to the Chilean case, which requires further
comparative as well as longitudinal research in the area. In any case,
teaching about the origins and characteristics of social inequality is
not part of the chilean school curriculum, calling for a more dedicated
consideration of these topics in areas such as citizensip education as
well as in history and social science education.

@@

The consideration of the school context opened some caveats in this
research. We found contrary evidence for the association between status,
school performance, and market justice preferences: high status and high
achievement at the school level are associated with less market justice
preferences. This is opposed to the rational-interest argument that
suggests that those doing better - both economically and academically-
would demand less redistribution \citep{runst_does_2018,bullock_education_2021,gonthier_parallel_2017} and therefore
are expected to challenge market mechanisms. A possible explanation is
that education contributes to forming a critical view of markets in
redistribution, nevertheless, further research in this area is needed to
assert the specific mechanisms that might be playing a role. All in all,
from this study we know that school contexts are not trivial in this
regard, calling for larger attention to school-age research on the
formation of inequality preferences and beliefs.

@@

\authorcontributions{Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: \getfirst{Juan Carlos}.\getfirst{Castillo}.:   Conceptualization, Supervision, Methodology, Writing
- review and
editing; \getfirst{Mauricio}.\getfirst{Salgado}.:   Conceptualization, Methodology, Writing
- review and
editing; \getfirst{Kevin}.\getfirst{Carrasco}.:   Methodology, formal
analysis, software, visualization, Writing - review and
editing; \getfirst{Andreas}.\getfirst{Laffert}.:   Methodology, formal
analysis, software, visualization, Writing - review and editing}

@@

\funding{The authors acknowledge the support of ANID/FONDECYT grant
1210847 and the Center for Conflict and Social Cohesion Studies (COES)
grant ANID/FONDAP 1513009.}

@@

\dataavailability{The original data presented in the study are openly
available in
\url{https://www.agenciaeducacion.cl/orientar/estudios/estudios-nacionales/}

Access to the data, code, and analysis presented in the study are openly available in \url{https://github.com/educacion-meritocracia/market-justice-preferences-school}

@@

\conflictsofinterest{The authors declare no conflict of interest.}

>>>>

% TODO: Add custom LaTeX header directives here
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\usepackage{threeparttable}
\usepackage{graphicx}
\includegraphics{pics/logo-mdpi.pdf}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\usepackage[]{natbib}
\bibliographystyle{plainnat}
\usepackage{bookmark}

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={The socialization of meritocracy and market justice preferences at school},
  pdfauthor={Equipo EDUMER},
  pdfkeywords={market
justice, meritocracy, socialization, family, schools, Chile},
  pdfcreator={LaTeX via pandoc},
  unicode=true,
	bookmarksopen={true},
	pdffitwindow=true, 
	colorlinks=true, 
	linkcolor=bluecite, 
	citecolor=bluecite, 
	urlcolor=bluecite, 
	hyperfootnotes=true, 
	pdfstartview={FitH},
	pdfpagemode= UseNone}

@@

% Author Orchid ID: enter ID or remove command
%\newcommand{\orcidauthorA}{0000-0000-0000-000X} % Add \orcidA{} behind the author's name
%\newcommand{\orcidauthorB}{0000-0000-0000-000X} % Add \orcidB{} behind the author's name

@@

\Author{Juan Carlos Castillo$^{aff-1}$, 
Mauricio Salgado$^{aff-2}$, 
Kevin Carrasco$^{aff-3}$~and~Andreas Laffert$^{aff-4}$
}
%\longauthorlist{yes}

% MDPI internal command: Authors, for metadata in PDF
\AuthorNames{
  Juan Carlos Castillo,
  Mauricio Salgado,
  Kevin Carrasco,
  Andreas Laffert}

% MDPI internal command: Authors, for citation in the left column
%\AuthorCitation{Lastname, F.; Lastname, F.; Lastname, F.}
%\usepackage{xstring}
\newcommand\getfirst[1]{\getfirstaux#1\relax} \def\getfirstaux#1#2\relax{#1}
\newcommand{\authorformatted}{Castillo, \getfirst{Juan Carlos}.; Salgado, \getfirst{Mauricio}.; Carrasco, \getfirst{Kevin}.; Laffert, \getfirst{Andreas}.}
\AuthorCitation{\authorformatted}

@@

% Affiliations / Addresses (Add [1] after \address if there is only one affiliation.)
\address {%
$^{aff-1}$ \quad , Department of Sociology, Universidad de Chile\\
$^{aff-2}$ \quad , Centro de Estudios Públicos, Chile\\
$^{aff-3}$ \quad , Center for Social Conflict and Cohesion Studies\\
$^{aff-4}$ \quad , Institute of Sociology, Universidad Católica de
Chile\\
}

@@

% Contact information of the corresponding author
%% see: https://quarto.org/docs/journals/authors.html
\corres{Correspondence: juancastillov@uchile.cl}

@@

Robert E. Lane proposed the underpinnings of the concept of market
justice, which he differentiated from political justice. For him, ``it
is the genius of the market to stimulate wants without at the same time
stimulating a sense of deserving more than one gets''
\citeyearpar[p.~384]{lane_market_1986}. Contrary to the evidence that
unequal distribution produces feelings of dissatisfaction, anger, and
resentment that might motivate forms of collective action
\citep{greitemeyer_subjective_2016, mishra_subjective_2015, smith_relative_2012, power_deprivationprotest_2018},
Lane pointed out that in market settings, social comparisons are more
likely to motivate increased effort rather than feelings of acute
injustice because individuals attribute outcomes to their actions. In
this sense, unequal levels of well-being would be, to some extent, a
function of their talents and efforts, instead of being based on
distributive principles that characterize welfare states, such as need
and equality (see \citep{wilson_role_2003}).

@@

Attending to the socialization within the family, the classic work of
Kohn showed that middle-class parents value the expression of internal
states and emotions, such as self-control, curiosity, happiness, and
consideration, while working-class parents promote deference, obedience,
and conformity to authority \citep{kohn_social_1963, kohn_class_1969}.
Although parents from all social backgrounds encourage individualism in
their children, this shared norm translates into different forms in high
and low social classes \citeyearpar{kohn_class_1969}.
\citet{acemoglu_obedience_2021} claimed that the values families impart
to their children interact with social mobility. On the one hand,
children from privileged families are socialized to adopt a clear
conception of individualism that highlights their internal states,
independence, and idiosyncrasies. In contrast, children from
disadvantaged families are socialized to support a more balanced view of
individualism that considers personal characteristics as resources to
overcome collective impediments on the path to upward mobility
\citep{iacoviello_collectivism_2019}. In the same line,
\citet{almas_fairness_2017} found that adolescents from
low-socioeconomic-status families are likelier to have an egalitarian
fairness view and consider an equal distribution as fair in a situation
with unequal merits. Taking this into account, we believe that there
will be differences in the socialization of values according to
socioeconomic differences in families that could influence market
justice preferences, which leads us to the following hypothesis:

@@

\includegraphics{paper_files/figure-pdf/fig-marketjustice-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-meritocracy-1.pdf}

@@

\includegraphics{paper_files/figure-pdf/fig-bivariate-1.pdf}

@@

\resizebox{\linewidth}{!}{
\begin{tabu} to \linewidth {>{\raggedright\arraybackslash}p{4cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}>{\centering\arraybackslash}p{1.5cm}}
\toprule
\multicolumn{1}{c}{\textbf{ }} & \multicolumn{5}{c}{\textbf{Market Justice Preferences}} \\
\cmidrule(l{3pt}r{3pt}){2-6}
Variable & School talent & School effort & Social talent & Social effort & Deservingness\\
\midrule
\textbf{University or Postgraduate} & -0.034 & -0.012 & -0.073. & -0.079* & -0.118**\\
\cmidrule{1-6}
\textbf{More than 25 books} & -0.015 & -0.046 & -0.029 & -0.081** & -0.035\\
\cmidrule{1-6}
\textbf{School SES high} & 0.004 & 0.001 & -0.057. & -0.063* & -0.08**\\
\cmidrule{1-6}
\textbf{Simce Medium} & 0.001 & 0.032 & -0.037 & -0.041 & -0.082*\\
\cmidrule{1-6}
\textbf{Simce High} & -0.016 & 0.008 & -0.043 & -0.063. & -0.087*\\
\bottomrule
\multicolumn{6}{l}{\textsuperscript{} Note: ***p < 0.001, **p < 0.01, *p < 0.05. Source: own elaboration based on First Study of Civic}\\
\multicolumn{6}{l}{Education 2017.}\\
\end{tabu}}

@@

\includegraphics{paper_files/figure-pdf/fig-interaction-1.pdf}

@@

Our research aimed to examine the relationship between the perception of
meritocracy and the justification of market justice preferences among
eighth-grade students in Chile. From our knowledge this is the first
study addressing market justice preferences at school level, finding
that students exhibit large preferences for market justice (about one
third agree or strongly agree with them), a share that is quite larger
when compared with evidence in adult population. Perception of
meritocracy is also high, particularly when it comes to the reward of
effort, with an striking eighty six percent of agreement. Regarding
these descriptive indicators, a a first message coming from this study
is that there is something particular in school-age when it comes to
perceptions and preferences about inequality. This can relate to
development aspects
\citep[\citet{rizzo_children_2020}]{kim_socioeconomic_2020} but it could
be something specific to the Chilean case, which requires further
comparative as well as longitudinal research in the area. In any case,
teaching about the origins and characteristics of social inequality is
not part of the chilean school curriculum, calling for a more dedicated
consideration of these topics in areas such as citizensip education as
well as in history and social science education.

@@

The consideration of the school context opened some caveats in this
research. We found contrary evidence for the association between status,
school performance, and market justice preferences: high status and high
achievement at the school level are associated with less market justice
preferences. This is opposed to the rational-interest argument that
suggests that those doing better - both economically and academically-
would demand less redistribution \citep[\citet{runst_does_2018},
\citet{bullock_education_2021}]{gonthier_parallel_2017} and therefore
are expected to challenge market mechanisms. A possible explanation is
that education contributes to forming a critical view of markets in
redistribution, nevertheless, further research in this area is needed to
assert the specific mechanisms that might be playing a role. All in all,
from this study we know that school contexts are not trivial in this
regard, calling for larger attention to school-age research on the
formation of inequality preferences and beliefs.

@@

@@

@@

@@

### 2024-08-05 paper.qmd
Actualización paper anonimizado en .tex y orden de repo

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  hikmah-pdf: 
    indent: false
  hikmah-manuscript-pdf: 
    indent: false
author:
  - name: Juan Carlos Castillo
    email: juancastillov@uchile.cl
    orcid: 0000-0003-1265-7854
    affiliations: 
        name: Department of Sociology, Universidad de Chile
        country: Chile
    attributes:
      corresponding: true
  - name: Mauricio Salgado
    email: msalgado@unab.cl
    orcid: 0000-0002-8108-6154
    affiliations: 
        name: Centro de Estudios Públicos, Chile
        country: Chile
  - name: Kevin Carrasco
    email: kevin.carrasco@ug.uchile.cl
    orcid: 0000-0002-9348-8361
    affiliations: 
        name: Center for Social Conflict and Cohesion Studies
        country: Chile
  - name: Andreas Laffert
    email: alaffertt@estudiante.uc.cl
    orcid: 0000-0002-9008-2454
    affiliations: 
        name: Institute of Sociology, Universidad Católica de Chile
        country: Chile
abstract: |
  Previous research has shown that schools justify student performance differences based on meritocratic ideals. However, the link between school meritocracy and students’ justification of inequalities has received little attention. This paper examines how school and family socialization influence students’ beliefs about meritocracy and market-based justifications for access to health, pensions, and education in Chile. Using data from the First Study of Civic Education in Chile, we employ a multilevel modeling strategy to analyze 5,047 students and parents nested within 231 schools. Our findings reveal that a significant proportion of Chilean students agree with market justice principles, at a higher rate than adults. Most students endorse meritocratic views, particularly the reward of effort. Multilevel estimations show strong associations between meritocratic beliefs and market justice preferences: students who believe in the reward of effort and talent are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with better achievement. Additionally, the conditional influence of meritocracy beliefs weakens in schools with higher socioeconomic status and performance. These results suggest that meritocratic beliefs and market justice preferences begin early and are influenced by the school environment.
keywords: 
  - market justice 
  - meritocracy
  - socialization
  - family
  - schools
  - Chile
bibliography: "input/bib/just-ineq-merit.bib"
csl: "input/bib/multidisciplinary-digital-publishing-institute.csl"
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"
papersize: letter
colorlinks: true
fig-height: 4
fig-width: 7.5
header-includes:
  - \usepackage{threeparttable}
---

This document was last modified at {{< modified >}}
and it was last rendered at {{< now >}}.

>>>>

---
title: "The socialization of meritocracy and market justice preferences at school"
format:
  mdpi-pdf:
    keep-tex: true
mdpi_journal: behavsci
article_type: article
article_status: submit
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    role:
      - Conceptualization
      - Supervision
      - Methodology
      - Writing - review and editing
    affiliations:
      - name: "Universidad de Chile"
        department: Department of Sociology
        id: 1
        country: Chile
  - name: Mauricio Salgado
    orcid: 0000-0002-8108-6154
    role:
      - Conceptualization
      - Methodology
      - Writing - review and editing    
    affiliations:
      - name: Centro de Estudios Públicos
        id: 2
        department: CEP
        country: Chile
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    role:
      - Methodology
      - formal analysis
      - software
      - visualization
      - Writing - review and editing
    affiliations:
      - name: Center for Social Conflict and Cohesion Studies
        id: 3
        department: COES
        country: Chile
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    role:
      - Methodology
      - formal analysis
      - software
      - visualization
      - Writing - review and editing  
    affiliations:
      - name: Universidad Católica de Chile
        id: 4
        department: Institute of Sociology
        country: Chile
fundingstatement: | 
  The authors acknowledge the support of ANID/FONDECYT grant 1210847 and the Center for Conflict and Social Cohesion Studies (COES) grant ANID/FONDAP 1513009.
dataavailability: |
  The original data presented in the study are openly available in <https://www.agenciaeducacion.cl/orientar/estudios/estudios-nacionales/>
conflictsofinterest: |
  The authors declare no conflict of interest.
abstract: |
  Previous research has shown that schools justify student performance differences based on meritocratic ideals. However, the link between school meritocracy and students’ justification of inequalities has received little attention. This paper examines how school and family socialization influence students’ beliefs about meritocracy and market-based justifications for access to health, pensions, and education in Chile. Using data from the First Study of Civic Education in Chile, we employ a multilevel modeling strategy to analyze 5,047 students and parents nested within 231 schools. Our findings reveal that a significant proportion of Chilean students agree with market justice principles, at a higher rate than adults. Most students endorse meritocratic views, particularly the reward of effort. Multilevel estimations show strong associations between meritocratic beliefs and market justice preferences: students who believe in the reward of effort and talent are more likely to justify inequalities based on financial capacity. At the school level, market justice preferences are higher in high-status schools but lower in schools with better achievement. Additionally, the conditional influence of meritocracy beliefs weakens in schools with higher socioeconomic status and performance. These results suggest that meritocratic beliefs and market justice preferences begin early and are influenced by the school environment.
keywords: [market justice, Meritocracy, socialization, family, schools, Chile]
bibliography: "input/bib/just-ineq-merit.bib"
---
